Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2015-10-07 paper.tex
label must go last in table or referencing doesn't work right
remove done todos
provide more definition of fleet and continuous

TODO: add table titles
TODO: adjust figure fonts to meet NT guidlines

@@

    \item Groups of facilities (usually all facilities of a given type) are
        treated as a single, aggregate entity.

    \item Resource flows are often (approximately) continuous (limited by time
        step discretization).

@@

\begin{table}[h]
    \label{tab:reactors}
    \caption{Reactor Facility Parameters}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & SFR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
\end{table}

@@

\begin{table}
    \label{tab:invar}
    \captionsetup{justification=centering}
    \caption{Reactor Parameter Invariants}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
                                                                             & LWR      & SFR      \\
        \hline                       
                                                                             &          &          \\
        Discharge Rate ($\frac{\text{kg} \cdot \text{HM}}{\text{month}}$)           & 1642.5   & 535      \\
                                                                             &          &          \\
        Burnup  ($\frac{\text{MWe} \cdot \text{month}}{\text{kg} \cdot \text{HM}}$) & 0.547945 & 0.672897 \\
                                                                             &          &          \\
        Effective Power  ($\text{MWe}$)                                      & 900      & 360      \\
                                                                             &          &          \\
        Core Size  ($\text{kg} \cdot \text{HM}$)                                    & 88695    & 40125    \\
                                                                             &          &          \\
        \hline                       
    \end{tabular}
\end{table}

\begin{table*}
    \label{tab:reactor-detail}
    \centering
    \captionsetup{justification=centering}
    \caption{Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
\end{table*}

@@

There are various low level phenomena that have potential to affect the larger
outcome of simulations with respect to facility discretization and time step modeling
decisions. Several effects are described below that arise from these
different modeling choices.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/sync-cycle.eps}
    \caption[The cycle staggering effect]{
        In this scenario, reactors are deployed annualy with a refueling cycle
        length (including outage) of 18 months.  So all reactors deployed
        every 3rd year refuel together.  All initial 100 LWRs start out with
        their cycles synced as well.  While it may look like separate curves,
        generated power curve (purple) represents only a single time series with many discontinuities.
    }
    \label{fig:sync-cycle}
\end{figure}

Another difference between fleet-based and individual facility modeling
involves fuel sharing.  Power production by a group of individual reactors for
a fixed amount of fuel shortage may vary depending on how available fuel is
distributed.  This is illustrated in Figure \ref{fig:fuel-sharing}.  A system
being short 3 batches might mean that one reactor has no batches and isn't
operating or that 3 reactors are each missing a single batch.  This is
referred to as the \emph{fuel sharing effect}.  Fleet reactors, by design,
have perfect fuel sharing - all fuel is distributed so that the minimum
possible amount of capacity is offline during shortages. The most correct
measure of fuel sharing inefficiency is the difference between the actual
generated power and the maximum amount of power that could have been generated
among all possible fuel distribution alternatives.  This is difficult to
actually compute. One possible approximation is to count the number of fresh
fuel batches distributed to reactors that ended up not having enough fuel to
operate (i.e. a full core) that time step.  This approximation assumes that
every "wasted" batch of fuel, could have been given to a reactor that only
needed one batch to operate.

@@

Figure \ref{fig:unfueled} shows the fuel shortage more explicitly for each of
the four cases. The individual reactor modeling cases result in more
cumulative outage than corresponding fleet-based cases. The quarterly time
step of cases QI and QF also make the outage significantly more severe. Cases MI
and QI exhibit approximately twice as much offline power as their corresponding
fleet based simulations in cases MF and QF.  The monthly time step simulations
also have roughly twice as much offline power as their corresponding quarterly
cases.

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} gives one way to approximate the \emph{fuel
sharing effect}.  Fleet reactor modeling by design exhibits perfect fuel
sharing by design (zero inefficiency) and so is not included in this figure.
Every batch that is given to a reactor that ends up not being able to start its
cycle at the scheduled time (due to fuel shortage) adds one to this cumulative
for each time step the cycle start is delayed.

@@

Quarterly time steps have less frequent but larger impulse material flows.
The larger in-flows, however, do not compensate for the larger withdrawals
because facilities do not know about incoming inventory when they resolve
their outflow for a particular time step.  This information lag effectively
requires the floating Pu inventory to be maintained at a higher level during
the shortage.  This can be clearly seen in the Figure \ref{fig:puinv} case QF plot which has a higher
standing inventory than case MF during the fuel shortage years.  The case QF Pu
inventory is unable to drop below approximately 100 tonnes, which indicates
out-flows to be roughly equal to the Pu flow into inventory each time step
during those years.  This can be confirmed by the in-flow and inventory curves
touching at those points.  Case MF Pu inventory is similarly limited at about
the 30 tonne level.  As expected, the case QF withdrawals are also three times
higher than the case MF withdrawals, exactly matching the time step duration ratio
between the two cases.

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feed-back effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepency between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF. This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

The horizontal striations visible in the monthly flows for case MI in Figure
\ref{fig:puinv} are a characteristic of the synchronization of reactor
refueling caused by the shortage.  They begin at the end of the fuel shortage
window when many reactors come online together and are resonances that the
inventory level jumps between.  These resonances gradually disappear as the
simulation progresses farther in time beyond the shortage and refueling cycles
become more staggered.

The in-flows shown for cases MF and QF show a periodic pulsing that begins
during the shortage in about year 120.  This pusling actually has nothing to
do with the shortage and is caused by the retirement of fast reactors.  The
first fast reactors are deployed in year 35 and have an 80 year lifetime.
When they retire, they discharge an entire core's worth of fuel rather than a
single batch.  After a 7-year cooling period, the first discharged full cores
begin to make it through the recycle loop in year 122.  Another interesting
feature visible in all four cases in Figure \ref{fig:puinv} is the minority
fraction of out-flow points that are higher than the bulk majority. These
higher out-flows occur once every 21 months when new reactors are built because
new reactors require a full core of fresh fuel rather than the single batch
needed when just refueling.

@@

Figure \ref{fig:puinv-compare} shows the same inventory curves from Figure
\ref{fig:puinv} superimposed. Perhaps counter-intuitively, lower separated plutonium
inventory during the shortage generally indicates fewer unfueled reactors.
Lower inventory levels mean available Pu is being utilized more efficiently
rather than sitting idle.  Cases QI and QF with their longer time steps show
higher inventories indicating a more severe shortage.  Post shortage, cases QI and QF
also fall a bit behind with respect to building up Pu stocks due to the delayed
contributions of more unfueled reactors to the separated Pu pool.  The
reactors in all the scenarios request fresh fuel only when they need it. While
this is an unrealistic behavior for real-world facilities, it is closer to a
globally optimum fuel management strategy.  Keeping on-hand fresh fuel batches
would also have the same effect of increasing idling plutonium quantities and
would make the fuel shortage worse overall.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/puinv-compare.eps}
    \caption{
        Separated Pu inventories for all four cases.  Differences in modeling
        choices are not particularly significant until about year 100 when
        fuel shortages begin.  These differences gradually decay away starting
        in year 140 after fuel shortages have ebbed.  Higher aggregate Pu
        inventory  during the shortage is correlated with both a longer shortage and
        a longer post-shortage recovery period.
    }
    \label{fig:puinv-compare}
\end{figure*}

>>>>

@@

    \item Groups of facilities (usually all facilities of a given type) are
        treated as a single, aggregate entity. Therefore, all facilities experience
        identical operational histories.

    \item Resource flows are modeled as continuous, within the limitations of time
        step discretization.  Therefore, material is acquired and discharge in
        uniform amounts at every time step.

@@

\begin{table}[h]
    \caption{Reactor Facility Parameters}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & SFR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \label{tab:reactors}
\end{table}

@@

\begin{table}
    \captionsetup{justification=centering}
    \caption{Reactor Parameter Invariants}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
                                                                             & LWR      & SFR      \\
        \hline                       
                                                                             &          &          \\
        Discharge Rate ($\frac{\text{kg} \cdot \text{HM}}{\text{month}}$)           & 1642.5   & 535      \\
                                                                             &          &          \\
        Burnup  ($\frac{\text{MWe} \cdot \text{month}}{\text{kg} \cdot \text{HM}}$) & 0.547945 & 0.672897 \\
                                                                             &          &          \\
        Effective Power  ($\text{MWe}$)                                      & 900      & 360      \\
                                                                             &          &          \\
        Core Size  ($\text{kg} \cdot \text{HM}$)                                    & 88695    & 40125    \\
                                                                             &          &          \\
        \hline                       
    \end{tabular}
    \label{tab:invar}
\end{table}

\begin{table*}
    \centering
    \captionsetup{justification=centering}
    \caption{Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \label{tab:reactor-detail}
\end{table*}

@@

There are various low level phenomena that have potential to affect the larger
outcome of simulations with respect to facility discretization and time step
modeling decisions. Several effects arising from these different modeling
choices are present in many fuel cycle simulators to varying degrees.  Some of
these effects are described in detail below.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/sync-cycle.eps}
    \caption[The cycle staggering effect]{
        In this scenario, reactors are deployed annualy with a refueling cycle
        length (including outage) of 18 months.  So all reactors deployed
        every 3rd year refuel together.  All initial 100 LWRs start out with
        their cycles synced as well.
    }
    \label{fig:sync-cycle}
\end{figure}

Another difference between fleet-based and individual facility modeling
involves fuel sharing.  Power production by a group of individual reactors for
a fixed amount of fuel shortage may vary depending on how available fuel is
distributed.  This is illustrated in Figure \ref{fig:fuel-sharing}.  A system
being short 3 batches might mean that one reactor has no batches and isn't
operating or that 3 reactors are each missing a single batch.  This is
referred to as the \emph{fuel sharing effect}.  Fleet reactors, by design,
have perfect fuel sharing - all fuel is distributed so that the minimum
possible amount of capacity is offline during shortages. The most correct
measure of fuel sharing inefficiency is the difference between the actual
generated power and the maximum amount of power that could have been generated
among all possible fuel distribution alternatives.  This is difficult to
actually compute. One possible approximation is to count the number of fresh
fuel batches distributed to reactors that ended up not having enough fuel to
operate (i.e. a full core) that time step.  This approximation assumes that
every "wasted" batch of fuel, could have been given to a reactor that only
needed one batch to operate.  Equation \ref{eq:badshare} describes such an
approximation.

\begin{equation}
    N_{wasted}(t) = \sum\limits_{r \in R_t} \left(H[t-S_{sched}(t,r)] \cdot [1-O(t,r)]  \sum\limits_{t'=F_{prev}(t,r)}^{t} N_b(t',r) \right)
    \label{eq:badshare}
\end{equation}

where $R_t$ is the set of all reactors, $H$ is the Heaviside function,
$S_{sched}(t,r)$ is the original scheduled startup time of reactor $r$ after
the most recent refueling outage that started before $t$, $F_{prev}(t,r)$ is
the start of the most recent refueling outage for reactor $r$ before or on
time $t$, $O(t,r)$ is 1 if reactor $r$ is producing power at time $t$ and zero
otherwise, and $N_b(t,r)$ is the number of new fuel batches received by
reactor $r$ since time $t$.

@@

For the fleet reactors, fuel shortage is exactly the
difference between generated power and installed power capacity. Measuring the
fuel shortage for individual reactor cases, however, requires careful
accounting of the difference between reactors that are offline for a normal
refueling outage and reactors that are offline because they have insufficient
fuel. Equation \ref{eq:shortage} shows how this is calculated for cases MI and
QI.

\begin{equation}
    P_{outage}(t) = \sum\limits_{r \in R_t} C_r \cdot H[t-S_{sched}(t,r)] \cdot [1-O(t,r)]
    \label{eq:shortage}
\end{equation}

where $R_t$ is the set of all reactors, $C_r$ is the power capacity of reactor
$r$, $H$ is the Heaviside function, $S_{sched}(t,r)$ is the original scheduled
startup time of reactor $r$ after the most recent refueling outage that
started before $t$, and $O(t,r)$ is 1 if reactor $r$ is producing power at
time $t$ and zero otherwise.

Figure \ref{fig:unfueled} shows the fuel shortage more explicitly for each of
the four cases with the case MI and QI results being the cumulative version of
Equation \ref{eq:shortage}.  The individual reactor modeling cases result in
more cumulative outage than corresponding fleet-based cases. The quarterly
time step of cases QI and QF also make the outage significantly more severe.
Cases MI and QI exhibit approximately twice as much offline power as their
corresponding fleet based simulations in cases MF and QF.  The monthly time
step simulations also have roughly twice as much offline power as their
corresponding quarterly cases.

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} provides one way to see the \emph{fuel sharing effect}
showing wasted batch-months computed using Equation \ref{eq:badshare}.  Fleet
reactor modeling by design exhibits perfect fuel sharing by design (zero
inefficiency) and so is not included in this figure.  Every batch that is
given to a reactor that ends up not being able to start its cycle at the
scheduled time (due to fuel shortage) adds one to this cumulative for each
time step the cycle start is delayed.

@@

Quarterly time steps have less frequent but larger impulse material flows.
The larger in-flows, however, do not compensate for the larger withdrawals
because facilities do not know about incoming inventory when they resolve
their outflow for a particular time step.  This information lag effectively
requires the floating Pu inventory to be maintained at a higher level during
the shortage.  This can be clearly seen in Figures \ref{fig:puinv} and \ref{fig:puinv-compare} where case QF has a higher
standing inventory than case MF during the fuel shortage years.  The case QF Pu
inventory is unable to drop below approximately 100 tonnes, which indicates
out-flows to be roughly equal to the Pu flow into inventory each time step
during those years.  This can be confirmed by the in-flow and inventory curves
touching at those points.  Case MF Pu inventory is similarly limited at about
the 30 tonne level.  As expected, the case QF withdrawals are also three times
higher than the case MF withdrawals, exactly matching the time step duration ratio
between the two cases.

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feed-back effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepency between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF (shown most clearly in Figure \ref{fig:puinv-compare-zoom}). This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

The horizontal striations visible in the monthly flows for case MI in Figure
\ref{fig:puinv} are a characteristic of the synchronization of reactor
refueling caused by the shortage.  A close-up view of these can be seen in
Figure \ref{fig:puinv-MI}. They begin at the end of the fuel shortage window
when many reactors come online together in groups and are resonances that the
inventory level jumps between.  These resonances gradually disappear as the
simulation progresses farther in time beyond the shortage and refueling cycles
become more staggered.

The in-flows shown for cases MF and QF show a periodic pulsing that begins
during the shortage in about year 120 and is shown in greater detail in Figure \ref{fig:puinv-MF}. This pulsing actually has nothing to
do with the shortage and is caused by the retirement of fast reactors.  The
first fast reactors are deployed in year 35 and have an 80 year lifetime.
When they retire, they discharge an entire core's worth of fuel rather than a
single batch.  After a 7-year cooling period, the first discharged full cores
begin to make it through the recycle loop in year 122.  Another interesting
feature visible in all four cases in Figure \ref{fig:puinv} is the minority
fraction of out-flow points that are higher than the bulk majority. These
higher out-flows occur once every 21 months when new reactors are built because
new reactors require a full core of fresh fuel rather than the single batch
needed when just refueling.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/puinv-MI.eps}
    \caption[Separated Pu inventory and flow: Case MI Zoom]{
        Separated Pu inventory and flows for case MI.  At locations where the
        outflow dots lie on/near the inventory level, fuel shortages are
        taking place.  This begins around year 105 and continues until about
        year 140 where Pu withdrawals become lower than standing inventory.
        The horizontal bands in out-flows starting around year 140 are caused
        by many previously offline reactors coming online together in groups.
        These fuel inventory level resonances gradually disappear as the
        simulation progresses.
    }
    \label{fig:puinv-MI}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/puinv-MF.eps}
    \caption[Separated Pu inventory and flow: Case MF Zoom]{
        Separated Pu inventory and flows for case MF.  At locations where the
        outflow dots meet the inventory level, fuel shortages are taking
        place.  This begins around year 105 and continues until about year 140
        where Pu withdrawals become lower than standing inventory.  The
        periodic pulsing that begins just after year 120 is caused by the
        first fast reactor retirements; the first fast reactors deployed starting in
        year 35 retire annually after their 80 year lifetime and discharge an entire core's
        worth of fuel rather than the usual single batch for refueling.
    }
    \label{fig:puinv-MF}
\end{figure*}

Figure \ref{fig:puinv-compare} shows the same inventory curves from Figure
\ref{fig:puinv} superimposed. Perhaps counter-intuitively, lower separated
plutonium inventory during the shortage generally indicates fewer unfueled
reactors.  Lower inventory levels mean available Pu is being utilized more
efficiently rather than sitting idle.  Cases QI and QF with their longer time
steps show higher inventories indicating a more severe shortage.  Cases QI and
QF take longer to recover from the shortage and fall a bit behind with respect
to building up Pu stocks due to the delayed contributions of more unfueled
reactors to the separated Pu pool.  This can be seen most clearly in Figure
\ref{fig:puinv-compare-zoom}.  Cases MI and QF recover Pu inventory
post-shortage at about the same time and rate - this is consistent with their
approximately equivalent cumulative offline power capacity curves in Figure
\ref{fig:unfueled}.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/puinv-compare.eps}
    \caption{
        Separated Pu inventories for all four cases.  Differences in modeling
        choices are not particularly significant until about year 100 when
        fuel shortages begin.  These differences gradually decay away starting
        in year 140 after fuel shortages have ebbed.
    }
    \label{fig:puinv-compare}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/puinv-compare-zoom.eps}
    \caption{
        Separated Pu inventories for all four cases zoomed in on the fuel
        shortage years.  Higher aggregate Pu inventory  during the shortage is
        correlated with both a longer shortage and a slower post-shortage
        recovery of Pu inventory.
    }
    \label{fig:puinv-compare-zoom}
\end{figure*}

The reactors in all the scenarios request fresh fuel only when they need it.
While this is an unrealistic behavior for real-world facilities, it is closer
to a globally optimum fuel management strategy.  Keeping on-hand fresh fuel
batches would also have the same effect of increasing idling plutonium
quantities and would make the fuel shortage worse overall.
